---
title: "Differential detection analysis"
author:
- name: Jeroen Gilis
  affiliation: 
  - &one Applied Mathematics, Computer science and Statistics, Ghent University, Ghent, Belgium
  - &two Bioinformatics Institute, Ghent University, Ghent, Belgium
  - Data Mining and Modeling for Biomedicine, VIB Flemish Institute for Biotechnology, Ghent, Belgium
- name: Helena L Crowell
  affiliation: 
  - National Center for Genomic Analysis (CNAG), Barcelona, Spain 
- name: Davide Risso
  affiliation: 
  - Department of Statistical Sciences, University of Padova, Padova, Italy
  - Padua Center for Network Medicine, University of Padova, Padova, Italy
- name: Lieven Clement
  affiliation: 
  - *one
  - *two
package: "`r BiocStyle::pkg_ver('muscat')`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{"3. Differential detection"}
  %\VignettePackage{muscat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "`r file.path(system.file('extdata', package='muscat'), 'refs.bib')`"
abstract: > 
  <p> he found himself transformed in his bed into a gigantic insect...
---

<style type="text/css">
.smaller {
  font-size: 10px
}
</style>

```{r cache, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(BiocStyle)
```

*** 

Based on @Gilis2023

> Gilis J, Perin L, Malfait M, Van den Berge K,  
Assefa AT, Verbist B, Risso D, and Clement L:  
Differential detection workflows for  
multi-sample single-cell RNA-seq data.  
*bioRxiv* (2023). [DOI: 10.1101/2023.12.17.572043](https://doi.org/10.1101/2023.12.17.572043)

# Load packages {-}

```{r load-libs, message=FALSE,  warning=FALSE}
library(dplyr)
library(purrr)
library(tidyr)
library(scater)
library(muscat)
library(ggplot2)
library(patchwork)
```

# Introduction

# Setup

We will use the same data as in the differential state (DS) analyses described in an independent vignette, namely, scRNA-seq data acquired on PBMCs from 8 patients before and after IFN-$\beta$ treatment.

```{r load-data, message=FALSE}
library(ExperimentHub)
eh <- ExperimentHub()
query(eh, "Kang")
(sce <- eh[["EH2259"]])
```

We further apply some minimal filtering to remove low-quality genes and cells, and use `prepSCE()` to standardize cell metadata such that slots specifying cluster (`cell`), sample (`stim`+`ind`), and group (`stim`) identifiers conform with the `muscat` framework:

```{r prep-data}
sce <- sce[rowSums(counts(sce) > 0) > 0, ]
qc <- perCellQCMetrics(sce)
sce <- sce[, !isOutlier(qc$detected, nmads=2, log=TRUE)]
sce <- sce[rowSums(counts(sce) > 1) >= 10, ]
sce$id <- paste0(sce$stim, sce$ind)
sce <- prepSCE(sce, "cell", "id", "stim")
table(sce$cluster_id, sce$group_id)
table(sce$sample_id)
```

## Aggregation 

```{r pbs-det}
pb_sum <- aggregateData(sce,
    assay="counts", fun="sum",
    by=c("cluster_id", "sample_id"))
pb_det <- aggregateData(sce,
    assay="counts", fun="num.detected",
    by=c("cluster_id", "sample_id"))
t(head(assay(pb_det)))
```

@Qiu2020 demonstrated that binarizing scRNA-seq counts generates expression profiles that still accurately reflect biological variation.
This finding was confirmed by @Bouland2021, who showed that the frequencies of zero counts capture biological variability, and further claimed that a binarized representation of the single-cell expression data allows for a more robust description of the relative abundance of transcripts than counts.

```{r pbs-mds, fig.width=8, fig.height=4, fig.cap="Pseudobulk-level multidimensional scaling (MDS) plot based on (A) sum of counts and (B) sum of binarized counts (i.e., counting the number of detected features) in each cluster-sample."}
pbMDS(pb_sum) + ggtitle("Σ counts") +
pbMDS(pb_det) + ggtitle("# detected") +
plot_layout(guides="collect") +
plot_annotation(tag_levels="A") &
theme(legend.key.size=unit(0.5, "lines"))
```

## Analysis

Once we have assembled the pseudobulk data, we can test for DD using `pbDD()`. By default, a $\sim$`group_id` model is fit, and the last coefficient of the linear model is tested to be equal to zero.

```{r pbDD}
res_DD <- pbDD(pb_det, min_cells=0, filter="none", verbose=FALSE)
```

# Stagewise anaysis

```{r pbDS}
res_DS <- pbDS(pb_sum, min_cells=0, filter="none", verbose=FALSE)
```

```{r}
res <- stagewise_DS_DD(res_DS, res_DD, verbose=FALSE)
head(res[[1]][[1]]) # results for 1st cluster
```

## Comparison

```{r}
# for each approach, get adjusted p-values across clusters
ps <- map_depth(res, 2, \(df) {
    data.frame(
        df[, c("gene", "cluster_id")],
        p_adj.stagewise=df$p_adj,
        p_adj.DS=df$res_DS$p_adj.loc,
        p_adj.DD=df$res_DD$p_adj.loc)
}) |> 
    lapply(do.call, what=rbind) |>
    do.call(what=rbind) |>
    data.frame(row.names=NULL)
head(ps)
```

To get an overview of how different approaches compare, we can count the number of genes found differential in each cluster for a given FDR threshold:

```{r fig.width=12, fig.height=4}
# for each approach & cluster, count number 
# of genes falling below 5% FDR threshold
ns <- lapply(seq(0, 0.2, 0.005), \(th) {
    ps |>
        mutate(th=th) |>
        group_by(cluster_id, th) |>
        summarise(
            .groups="drop",
            across(starts_with("p_"), 
            \(.) sum(. < th, na.rm=TRUE)))
}) |> 
    do.call(what=rbind) |>
    pivot_longer(starts_with("p_"))
ggplot(ns, aes(th, value, col=name)) + 
    geom_line(linewidth=0.8, key_glyph="point") +
    geom_vline(xintercept=0.05, lty=2, linewidth=0.4) +
    guides(col=guide_legend(NULL, override.aes=list(size=3))) +
    labs(x="FDR threshold", y="number of significantly\ndifferential genes") +
    facet_wrap(~cluster_id, scales="free_y", nrow=2) + 
    theme_bw() + theme(
        panel.grid.minor=element_blank(),
        legend.key.size=unit(0.5, "lines"))
```

We can further identify which hits are shared between or unique to a given approach.
In the example below, for instance, the vast majority of hits is common to all approaches, many hits are shared between DD and stagewise testing, and only few genes are specific to any one approach:

```{r upset, fig.width = 5, fig.height = 3, fig.cap = "Upset plot of differential findings (FDR < 0.05) across DS, DD, and stagewise analysis for an exemplary cluster; shown are the 50 most frequent interactions."}
# subset adjuster p-values for cluster of interest
qs <- ps[grep("CD4", ps$cluster_id), grep("p_", names(ps))]
# for each approach, extract genes at 5% FDR threshold
gs <- apply(qs, 2, \(.) ps$gene[. < 0.05])
# visualize set intersections between approaches
UpSetR::upset(UpSetR::fromList(gs), order.by="freq")
```

```{r}
# extract genes unique to stagewise testing
sw <- grep("stagewise", names(gs))
setdiff(gs[[sw]], unlist(gs[-sw]))
```

# Session info {- .smaller}

```{r session-info}
sessionInfo()
```

# References